Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-07-07 paper/paper.tex
Camera ready
Font
minors

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$, for all morphisms
$f : X \to Y$. Intuitively, a natural transformation is akin to a polymorphic
function transforming the structure of a functor into the structure of another
functor without assuming what is the type of data contained in them.  For
example, a program $f : \texttt{List }X \to \texttt{Maybe} X$ which returns
nothing if the list is empty and the head of the list otherwise can work for all
types $X$ uniformly because it needs not to inspect the data inside the list.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For an endofunctor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}. These are  arrows $f : X \to Y$ in the underlying category $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
if $F$ has an initial $F$-algebra then this is the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ where
$\InIso : F \mu F \to \mu F$ is the $F$-algebra  witnessing the
isomorphism $F\mu F \iso \mu F$, furthermore  $\InOp : \mu F \to F \mu F$ is the
inverse of $\InIso$. The uniqueness property of initial $F$-algebras states that for any other
$F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
by $\cata{\alpha_{X}} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
``\emph{catamorphism}'' or ``\emph{fold}'' satisfying:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
\end{equation}
As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.

>>>>

Given two functors $F, G : \CatC \to \CatD$, a \emph{natural transformation} is
is a family of morphisms $\phi_{X} : FX \to GX$ indexed
by the objects $X \in \Obj(\CatC)$ and such that it is \emph{natural} in $X$,
that is $ G(f) \cdot \phi_{X} = \phi_{Y} \cdot F(f)$, for all morphisms
$f : X \to Y$. Intuitively, a natural transformation is akin to a polymorphic
function transforming the structure of a functor into the structure of another
functor without assuming what is the type of data contained in them.  For
example, a program $\phi_{X} : \texttt{List }X \to \texttt{Maybe} X$, which returns
nothing if the list is empty and the head of the list otherwise, does not need to know the content of the list to perform its task, in other words, it can work for all
types $X$ \emph{uniformly}.

\subsection{Algebras and Catamorphisms}
\label{sec:algebras}
For an endofunctor $F : \CatC \to \CatC$ an $F$-\emph{algebra} is a pair $(X,a_{X})$
where $X$ is an object of the category called the \emph{carrier} of the algebra
and $a_{X}$ is an arrow of type $FX \to X$ called the \emph{structure map}.
The category of $F$-algebras, denoted by $F\Alg(\CatC)$, is the category where
objects are $F$-algebras and arrows $f : (X, a_{X}) \to (Y, a_{Y})$ are
\emph{$F$-algebra homomorphisms}. These are  arrows $f : X \to Y$ in the underlying category $\CatC$ such that they
respect the structure of the algebra, that is
$f \circ a_{X} = a_{Y} \circ F(f)$. The initial object in this category is
called the \emph{initial $F$-algebra}, that is the $F$-algebra which has a
unique $F$-algebra homomorphism into any other $F$-algebra. By Lambek's lemma
if $F$ has an initial $F$-algebra then this is the \emph{least fixed-point} for the
functor $F$ which we denote by $(\mu F, \InIso)$ where
$\InIso : F \mu F \to \mu F$ is the $F$-algebra  witnessing the
isomorphism $F\mu F \iso \mu F$, furthermore  $\InOp : \mu F \to F \mu F$ is the
inverse of $\InIso$. The uniqueness property of initial $F$-algebras states that for any other
$F$-algebra $(X, a_{X})$ there exists a unique $F$-algebra homomorphism, denoted
by $\cata{\alpha_{X}} : (\mu F, \InIso) \to (X, a_{X})$ and pronounced
``\emph{catamorphism}'' or ``\emph{fold}''. This is formally stated as follows:
\begin{equation}
  \label{eq:cata:unique}
  f = \cata{a_{X}} \iff f = a_{X} \circ F(f) \circ \InOp
\end{equation}
As a result of the uniqueness property we can derive the fusion
law. For all $F$-algebra homomorphisms $f : (X, a_{X}) \to (Y, a_{Y})$ we have
\begin{equation}
  \label{eq:cata:fusion}
  f \circ \cata{a_{X}} = \cata{a_{Y}}
\end{equation}
which means that the composition of a program $f$ with a catamorphism recursing
once over the data structure is the same as performing that recursion once using
the algebra $a_{Y}$ instead of $a_{X}$. This is a useful result for program
optimisation as we shall see.
